Supplementary Materials for cILR: Taxonomic enrichment analysis with isometric log ratios

### 2021-09-07 manuscript_plos.tex
Fixing issues Annie talked about

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Research in human associated microbiomes often involves the analysis of taxonomic count tables generated via high-throughput sequencing. It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional. An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets. Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis. Unfortunately, there is a lack of sophisticated set-based analysis methods specific to microbiome data, where current practice often employs abundance summation as a technique for aggregation. This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias. Here, we attempt to fill this gap with a new single sample taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature. Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement. We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values. Herein we demonstrate using both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios. Additionally, it provides informative scores that can be inputs to downstream differential abundance and prediction tasks.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing. However, microbiome data is sparse and high dimensional which prevents the application of standard statistical techniques. One approach to address this problem is to perform analyses at the level of taxon sets. Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability. Unfortunately, there is not a lot of research in developing new set-based tools for microbiome data specifically, given that compared to other 'omics data types microbiome data is strictly compositional. We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis. Our scores can be used for statistical inference, and as inputs to other downstream analyses such as differential abundance and prediction models. We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.

@@

Second, the number of reads obtained is constrained by the sequencing instrument at an arbitrary limit, and applied inconsistently across samples, resulting in a different number of total read counts per sample. Many normalization methods \cite{weiss2017} have been proposed to address these issues, including cross-applying methods from the gene expression literature \cite{mcmurdie2014}. However, these methods rely on assumptions specific to the original bulk RNA-seq data sets such as the presence of housekeeping genes with consistent expression levels \cite{love2014}, which might not be not true in the context of microbiome relative abundance data \cite{quinn2019,quinn2018b}. As such, microbiome data is strictly compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another. Consequently, log-ratio transformations from the compositional data literature are often utilized \cite{aitchison1999}.

Third, the data is highly zero-inflated, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment). Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015}, or applying zero-inflated models \cite{chen2018, kaul2017}. Newer methods developed recently have focused on understanding the different types of zeros in the data, providing more sophisticated heuristics around when pseudocounts can be utilized \cite{kaul2017a}.

Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}. Aggregated sets can be less sparse than their constituent elements, and testing on a smaller number of variables can reduce the multiple testing burden, thereby increasing power and reproducibility. Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations. There exists a diverse set of available methods developed to perform such analyses. More traditional set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest \cite{goeman2007}. Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values. The most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size. Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples. Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample level and are more akin to a transformation. This strategy allows for the flexible incorporation of different statistical techniques downstream, such as prediction models, as well as for visualization purposes in ordination plots.

In microbiome research, even though no explicit enrichment analysis is performed, standard practice often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum by simple summation of abundances \cite{mclaren2019}. Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, it increases bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}. As such, there is a need for microbiome researchers to adopt more robust sample-level set enrichment methods. Unfortunately, limited work has been done to extend existing methods to be more specific to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do provide researchers tools to perform enrichment testing with curated taxon sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used.

Here, we developed a novel method that generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}. We leverage the concept of the $Q_1$ competitive hypothesis presented in Tian et al. \cite{tian2005}, which tests the null that the value of variables within a specific set is equal to the value of measured variables not in the set. The competitive null hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables. We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a multiplicative aggregation method that addresses the downsides of the naive summation-based method presented above \cite{mclaren2019, silverman2017}. The resulting method, titled competitive isometric log-ratio (cILR), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing. These scores can then act as inputs to differential abundance and predictive modeling tasks downstream.

In the following sections, we provide the formulation of cILR and discuss some statistical properties. We illustrate significance testing at the sample level using cILR and evaluate type I error and power under different simulation scenarios and real data applications. We assess the informativeness of cILR generated scores, and evaluate how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis. We compare the performance of cILR in these respective tasks against standard microbiome data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods. An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).

@@

    \item \textbf{Calculate finalized cILR scores with respect to the empirical null}. Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution. Valid p-values can be calculated by subtracting $\mathbf{E}$ from 1. 

@@

This assumption is also supported by preliminary simulation studies (detailed below). In panel A of Fig~\ref{fig:1}, we simulated microbiome count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness. It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation. Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis). The effect is most dramatic under both high inter-taxa correlation and sparsity. This heterogeneity further supports the decision to estimate an empirical null distribution, similar to Efron \cite{efron2004}.

@@

\noindent \textbf{Differential Abundance Analysis}: To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition. Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null. Power is defined similarly, but instead under the global alternate hypothesis. Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error. A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3). Due to the compositional nature of microbiome data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets. As such, we applied a compensation procedure as described in Hawinkel et al. \cite{hawinkel2019} to ensure the validity of simulation results. All sample sizes were set at 2,000.

@@

\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of cILR in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}. We utilized this data set following the approach outlined in Calagaro et al. \cite{calgaro2020}. This data set is special because it is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}. Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false positive rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment. Microbial tropism annotation at the genus level was from Beghini et al. \cite{beghini2019} and was downloaded directly from the GitHub repository associated with Calagaro et al. \cite{matteocalgaro2020}.

@@

Unfortunately, cILR did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}. This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation. However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and cILR/CLR. Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity. Additionally, we did not observe any performance difference between the different variations of cILR.

@@

\subsubsection*{Differential abundance analysis}
For differential abundance testing, we evaluated whether using cILR scores alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes. We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014} applied on data where taxa were aggregated using the naive sum method. We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome data \cite{mcmurdie2014}. Conversely, corncob is a newer method developed specifically for microbiome data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.

@@

\subsubsection*{Predictive models}
For disease prediction, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using cILR generated scores as inputs. Similar to our inference analysis, we compared cILR against both ssGSEA and GSVA. Additionally, we also evaluated cILR with the approach where counts of a set were aggregated using sums and then centered log-ratio transformed. This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}. Results indicated that cILR produces good performance values across both real data analysis and simulation scenarios. Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses. Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome data sets. This is consistent with our sample ranking analysis (Fig~\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts. However, cILR did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data.

@@

\subsection*{Limitations and future directions} 
There are various limitations to our evaluation of cILR. First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome data. There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation of prior studies. Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is not perfect. This is because the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut. As such, results from power analyses using this data set is only relative between the comparison methods. Finally, we assumed that taxa within a set are all equally associated with the outcome. This limits our ability to evaluate the performance of cILR when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.

@@

In addition, there are possible extensions cILR can we can consider to provide more flexibility across different data analysis scenarios in data analysis. First, cILR did not address the sparsity of microbiome data and relies on a pseudocount to ensure log operations are valid. We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}. Second, cILR also treated all taxa within the set as equally contributing to the set. Incorporation of taxa-specific weights can reduce the influence of outliers, such as rare or highly invariant taxa. Finally, curating sets based on apriori characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.

\section*{Conclusion}
Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomics data sets. However, limited work has been done developing set based methods specifically for microbiome relative abundance data. In this manuscript, we introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level. We demonstrated that our method can control for type I error for significance testing at the sample level, while generated scores are also valid inputs in downstream analyses, including disease prediction and differential abundance.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Research in human associated microbiomes often involves the analysis of taxonomic count tables generated via high-throughput sequencing. It is difficult to apply statistical tools as the data is high-dimensional, sparse, and strictly compositional. An approachable way to alleviate high-dimensionality and sparsity is to aggregate variables into pre-defined sets. Set-based analysis is ubiquitous in the genomics literature, and has demonstrable impact in improving interpretability and power of downstream analysis. Unfortunately, there is a lack of sophisticated set-based analysis methods specific to microbiome taxonomic data, where current practice often employs abundance summation as a technique for aggregation. This approach prevents comparison across sets of different sizes, does not preserve inter-sample distances, and amplifies protocol bias. Here, we attempt to fill this gap with a new single sample taxon set enrichment method based on the isometric log ratio transformation and the competitive null hypothesis commonly used in the enrichment analysis literature. Our approach, titled competitive isometric log ratio (cILR), generates sample-specific enrichment scores as the scaled log ratio of the subcomposition defined by taxa within a set and the subcomposition defined by its complement. We provide sample-level significance testing by estimating an empirical null distribution of our test statistic with valid p-values. Herein we demonstrate using both real data applications and simulations that cILR controls for type I error even under high sparsity and high inter-taxa correlation scenarios. Additionally, it provides informative scores that can be inputs to downstream differential abundance and prediction tasks.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
The study of human associated microbiomes relies on genomic surveys via high-throughput sequencing. However, microbiome taxonomic data is sparse and high dimensional which prevents the application of standard statistical techniques. One approach to address this problem is to perform analyses at the level of taxon sets. Set-based analysis has a long history in the genomics literature, with demonstrable impact in improving both power and interpretability. Unfortunately, there is not a lot of research in developing new set-based tools for microbiome taxonomic data specifically, given that compared to other 'omics data types microbiome taxonomic data is strictly compositional. We developed a new tool to generate taxon set enrichment scores at the sample level by combining the isometric log-ratio and the competitive null hypothesis. Our scores can be used for statistical inference, and as inputs to other downstream analyses such as differential abundance and prediction models. We demonstrate the performance of our method against competing approaches across both real data analyses and simulation studies.

@@

Second, the number of reads obtained is constrained by the sequencing instrument at an arbitrary limit, and is inconsistent across samples, resulting in a variable number of total read counts per sample. Many normalization methods \cite{weiss2017} have been proposed to address these issues, including cross-applying methods from the gene expression literature \cite{mcmurdie2014}. However, these methods rely on assumptions specific to the original bulk RNA-seq data sets such as the presence of housekeeping genes with consistent expression levels \cite{love2014}, which might not be not true in the context of microbiome relative abundance data \cite{quinn2019,quinn2018b}. As such, microbiome taxonomic data is strictly compositional \cite{gloor2017}, which means that the abundance of any taxa can only be interpreted relative to another. Consequently, log-ratio transformations from the compositional data literature are often utilized \cite{aitchison1999}.

Third, the data are highly zero-inflated, where there is a high number of both structural zeros (truly missing due to biological reasons) and sampling zeroes (due to limits of detection of the sequencing experiment). Researchers often dealt with these issues by imputing zero cells with a pseudocount \cite{kurtz2015}, or applying zero-inflated models \cite{chen2018, kaul2017}. Newer methods developed recently have focused on understanding the different types of zeros in the data, providing more sophisticated heuristics around when pseudocounts can be utilized \cite{kaul2017a}.

Even though the aforementioned problems are challenging, a very approachable method to address some of them is through set-based analysis, also termed gene set testing in the genomics literature \cite{khatri2012, goeman2007}. Aggregated sets are less sparse than their constituent elements, and testing on a smaller number of variables reduces the multiple testing burden, thereby increasing power and reproducibility. Through the usage of functionally informative sets defined apriori based on historical experiments (for example MSigDB \cite{subramanian2005}, and Gene Ontology \cite{ashburner2000}), gene set analyses also allows for more informative interpretations. There exists a diverse set of available methods developed to perform such analyses. More traditional set testing methods utilize the hypergeometric test to test for the overrepresentation of significant p-values for a set of interest \cite{goeman2007}. Unfortunately, these approaches are sensitive to the differential expression test and their generated p-values. The most widely used gene-set analysis method, GSEA \cite{subramanian2005}, instead uses a random-walk-like statistic through a ranking of genes based on a measure of association or effect size. Both of these methods generate enrichment scores and significance testing at the population level, incorporating information from all samples. Conversely, methods such as GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}, generate enrichment scores at the sample level and are more akin to a transformation. This strategy allows for the flexible incorporation of different statistical techniques downstream, such as prediction models, as well as for visualization purposes in ordination plots.

In microbiome research, even though no explicit enrichment analysis is performed, standard practice often involves aggregating taxa to higher Linnean classification levels such as genus, family, or phylum by simple summation of abundances \cite{mclaren2019}. Even though this allows for a reduction in the number of overall taxa (from thousands to only hundreds), there still exist three disadvantages: first, inter-sample distances are not preserved before and after aggregation \cite{egozcue2005}, second, it doesn't allow for enrichment testing and comparison across sets of different sizes, and third, it increases bias when taxa within the set have different efficiencies in how they are measured through sequencing \cite{mclaren2019}. As such, there is a need for microbiome researchers to adopt more robust sample-level set enrichment methods. Unfortunately, limited work has been done to extend existing methods to be more specific to microbiome relative abundance data. Some software suites, such as \emph{MicrobiomeAnalyst}, do offer tools to perform enrichment testing with curated taxon sets \cite{chong2020}. However, the approach used in \emph{MicrobiomeAnalyst} is a form of overrepresentation analysis at the population level and therefore similarly sensitive to the differential abundance approach used.

Here, we present a novel method that generates enrichment scores at the sample level similar to GSVA \cite{hanzelmann2013} and VAM \cite{frost2020a}. We leverage the concept of the $Q_1$ competitive hypothesis presented in Tian et al. \cite{tian2005}, which tests the null that the value of variables within a specific set is equal to the value of measured variables not in the set. The competitive null hypothesis is particularly useful in compositional data analysis, as it naturally assesses enrichment as a ratio between two sets of variables. We incorporated this insight with the isometric log-ratio transformation \cite{egozcue2003}, which allows for a multiplicative aggregation method that addresses the downsides of the naive summation-based method presented above \cite{mclaren2019, silverman2017}. The resulting method, titled competitive isometric log-ratio (cILR), is therefore unsupervised and can generate sample-specific enrichment scores with a well-defined null hypothesis that allows for significance testing. These scores can then act as inputs to differential abundance and predictive modeling tasks downstream.

In the following sections, we provide the formulation of cILR and discuss some statistical properties. We illustrate significance testing at the sample level using cILR and evaluate type I error and power under different simulation scenarios and real data applications. We assess the informativeness of cILR generated scores, and evaluate how it performs as part of downstream analyses, specifically predictive models and differential abundance analysis. We compare the performance of cILR in these respective tasks against standard microbiome taxonomic data analysis practices, as well as the existing GSVA \cite{hanzelmann2013} and ssGSEA \cite{barbie2009}, which are single sample methods. An R package implementation of this approach can be found on Github (\href{www.github.com/qpmnguyen/teaR}{qpmnguyen/teaR}).

@@

    \item \textbf{Calculate finalized cILR scores with respect to the empirical null}. Enrichment scores $\mathbf{E}_{i,k}$ are calculated as the cumulative distribution function (CDF) values or z-scores with respect to $\mathbf{P}_{emp}$ distribution. P-values can be calculated by subtracting $\mathbf{E}$ from 1. 

@@

This assumption is also supported by preliminary simulation studies (detailed below). In panel A of Fig~\ref{fig:1}, we simulated microbiome taxonomic count data under the global null across different data features and compute raw cILR scores and compute kurtosis and skewness. It can be seen that the characteristics of the null change depending on sparsity and inter-taxa correlation. Sparsity seems to drive the distribution to be more positively skewed while inter-taxa correlation encourages platykurtic (negative kurtosis). The effect is most dramatic under both high inter-taxa correlation and sparsity. This heterogeneity further supports the decision to estimate an empirical null distribution, similar to Efron \cite{efron2004}.

@@

\noindent \textbf{Differential Abundance Analysis}: To assess type I error rate and power for differential abundance testing task, we simulated data based on the schema above, and assessed differential abundance of 50 sets with 100 taxa per set across 20 replicates per simulation condition. Type I error is calculated as the number of differentially abundant sets over the total number of sets for each simulation under the global null. Power is defined similarly, but instead under the global alternate hypothesis. Estimates and confidence intervals for type I error and power are calculated as cross-replicate mean and standard error. A set is differentially abundant when all taxa within a set are differentially abundant with the same effect size. Across both analyses, we varied sparsity levels ($p = 0.2, 0.4, 0.6$), and inter-taxa correlation within the set ($\rho = 0, 0.2, 0.5$). Half of the sets are differentially abundant across case/control status with varying effect sizes (fold change of 1.5, 2, and 3). Due to the compositional nature of microbiome taxonomic data, simple inflation of raw counts would cause an artificial decrease in the abundance of the remaining un-inflated sets. As such, we applied a compensation procedure as described in Hawinkel et al. \cite{hawinkel2019} to ensure the validity of simulation results. All sample sizes were set at 2,000.

@@

\noindent \textbf{Single Sample Enrichment}: To assess the false discovery rate and true discovery rate of cILR in sample-level enrichment testing, we utilized the 16S rRNA gene sequencing of the oral microbiome at the gingival subsite from the Human Microbiome Project \cite{consortium2012, proctor2019}. We utilized this data set following the approach outlined in Calagaro et al. \cite{calgaro2020}. This data set is approximately labeled, where aerobic microbes are enriched in the supragingival subsite where the biofilm is exposed to the open air, while conversely anaerobic microbes thrive in the subgingival site \cite{thurnheer2016}. Here, we assessed the enrichment of aerobic microbes across all samples, we considered the false positive rate as the number of samples from the subgingival site with significant enrichment, and the true positive rate as the number of supragingival samples with significant enrichment. Microbial tropism annotation at the genus level was from Beghini et al. \cite{beghini2019} and was downloaded directly from the GitHub repository associated with Calagaro et al. \cite{matteocalgaro2020}.

@@

In this analysis, cILR unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}. This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation. However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and cILR/CLR. Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity. Additionally, we did not observe any performance difference between the different variations of cILR.

@@

\subsubsection*{Differential abundance analysis}
For differential abundance testing, we evaluated whether using cILR scores alongside a standard difference in means test (Welch's t-test and Wilcoxon rank sum test) is suitable to detect changes in abundance of a set of microbes. We compared cILR against two popular approaches: corncob \cite{martin2020} and DESeq2 \cite{love2014} applied on data where taxa were aggregated using the naive sum method. We chose DESeq2 because it is an older approach from the bulk RNA-seq literature that has strong support for usage in microbiome taxonomic data \cite{mcmurdie2014}. Conversely, corncob is a newer method developed specifically for microbiome taxonomic data sets, where taxonomic counts are modeled directly using a beta-binomial distribution instead of relying on normalization via size factor estimation like in DESeq2.

@@

\subsubsection*{Predictive models}
For disease prediction, we fitted a basic random forest model \cite{breiman2001} to predict continuous and binary outcomes using cILR generated scores as inputs. Similar to our inference analysis, we compared cILR against both ssGSEA and GSVA. Additionally, we also evaluated cILR with the approach where counts of a set were aggregated using sums and then centered log-ratio transformed. This is because CLR is considered standard practice in using microbiome variables as predictors for a model \cite{gloor2017}. Results indicated that cILR produces good performance values across both real data analysis and simulation scenarios. Since predictive models consider the effect of variables jointly (and in the case of random forest, consider interactions as well), good performance indicates that cILR scores can capture joint distribution of sets, enabling both uniset and multi-set type analyses. Comparatively, cILR generated scores outperformed other enrichment score methods (GSVA and ssGSEA), suggesting that it is more tailored for microbiome taxonomic data sets. This is consistent with our sample ranking analysis (Fig~\ref{fig:3}), where cILR scores are on average more informative when used to rank samples based on their propensity to have inflated counts. However, cILR did not outperform the CLR approach across our simulation studies, and only marginally performed better in the real data analysis with WGS data.

@@

\subsection*{Limitations and future directions} 
There are various limitations to our evaluation of cILR. First, our simulation analysis might not capture the appropriate data-generating distributions underlying microbiome taxonomic data. There is strong evidence to suggest that our zero-inflated negative binomial distribution is representative \cite{calgaro2020}, however other distributions such as the Dirichlet multinomial distribution \cite{wu2016} have been used in the evaluation of prior studies. Second, the usage of the gingival data set similar to \cite{calgaro2020} to assess power in differential abundance testing and single sample inference is not perfect. This is because the oxygen usage label of each microbe in the data set is only available at the genus level, and the difference in counts for obligate aerobes and anaerobes across the supragingival and subgingival sites might not be as clear cut. As such, results from power analyses using this data set is only relative between the comparison methods. Finally, we assumed that taxa within a set are all equally associated with the outcome. This limits our ability to evaluate the performance of cILR when only a small number of taxa within the set is associated with the outcome, or if there are variability in effect sizes or association direction of taxa within a set.

@@

In addition, there are possible extensions cILR can we can consider to provide more flexibility across different data analysis scenarios in data analysis. First, cILR did not address the sparsity of microbiome taxonomic data and relies on a pseudocount to ensure log operations are valid. We can address this by incorporating more sophisticated model-based zero-correction methods such as in \cite{martin-fernandez2012} or \cite{kaul2017a}. Second, cILR also treated all taxa within the set as equally contributing to the set. Incorporation of taxa-specific weights could reduce the influence of outliers, such as rare or highly invariant taxa. Finally, curating sets based on \emph{apriori} characteristics of microbes can allow for incorporating functional insights into microbiome-outcome analyses while also improving interpretability when compared to using taxonomic categories such as phylum or genus alone.

\section*{Conclusion}
Gene set testing, or pathway analysis is an important tool in the analysis of high-dimensional genomics data sets. However, limited work has been done developing set based methods specifically for microbiome relative abundance data. We introduced a new microbiome-specific method to generate set-based enrichment scores at the sample level. We demonstrated that our method can control for type I error for significance testing at the sample level, while generated scores are also valid inputs in downstream analyses, including disease prediction and differential abundance.
